### 2023-03-28 proofs/main.tex
Prove the UB lemma

@@

\begin{definition}
  foo

  $\begin{aligned}[t]
  [[us1 & us2]] &= \{ [[usEntry1 & usEntry2]] \mid [[usEntry1]] \in [[us1]],
  [[usEntry2]]  \in [[us2]], \text{s.t. } [[usEntry1]] \text{ matches with }
                                   [[usEntry2]] \}\\
                &\cup
                                     \{ [[usEntry1]] \mid [[usEntry1]] \in
                                     [[us1]], \text{ s.t. }
                                     \forall [[usEntry2]]  \in [[us2]],
                                     [[usEntry1]] \text{ does not match with }
                                                     [[usEntry2]] \}\\
        &\cup
          \{ [[usEntry2]] \mid [[usEntry2]] \in
          [[us2]], \text{ s.t. }
          \forall [[usEntry1]]  \in [[us1]],
          [[usEntry1]] \text{ does not match with }
          [[usEntry2]] \}\\
   \end{aligned}$
\end{definition}

@@

@@

\subsection{Variable Ordering}

\input{../_build/_gen/tex/ord-lemmas.tex}

\subsection{Normaliztaion}

\input{../_build/_gen/tex/norm-lemmas.tex}

>>>>

% G ⊢ N ≤1 M

\renewcommand{\ottdruleDOneNVarName}[0]{(Var$^-$$^{\subDOne}$)}
\renewcommand{\ottdruleDOneShiftUName}[0]{($\uparrow^{\subDOne}$)}
\renewcommand{\ottdruleDOneArrowName}[0]{($\rightarrow^{\subDOne}$)}
\renewcommand{\ottdruleDOneForallName}[0]{($\forall^{\subDOne}$)}

% G ⊢ P ≥1 Q
\renewcommand{\ottdruleDOnePVarName}[0]{(Var$^+$$^{\supDOne}$)}
\renewcommand{\ottdruleDOneShiftDName}[0]{($\downarrow^{\supDOne}$)}
\renewcommand{\ottdruleDOneExistsName}[0]{($\exists^{\supDOne}$)}

@@

\begin{definition}
  $\begin{aligned}[t]
  [[us1 & us2]] &= \{ [[usEntry1 & usEntry2]] \mid [[usEntry1]] \in [[us1]],
  [[usEntry2]]  \in [[us2]], \text{s.t. } [[usEntry1]] \text{ matches with }
                                   [[usEntry2]] \}\\
                &\cup
                                     \{ [[usEntry1]] \mid [[usEntry1]] \in
                                     [[us1]], \text{ s.t. }
                                     \forall [[usEntry2]]  \in [[us2]],
                                     [[usEntry1]] \text{ does not match with }
                                                     [[usEntry2]] \}\\
        &\cup
          \{ [[usEntry2]] \mid [[usEntry2]] \in
          [[us2]], \text{ s.t. }
          \forall [[usEntry1]]  \in [[us1]],
          [[usEntry1]] \text{ does not match with }
          [[usEntry2]] \}\\
   \end{aligned}$
\end{definition}

@@

\subsection{Substitution}
\input{../_build/_gen/tex/subst-lemmas.tex}

\subsection{Type well-formedness}
\input{../_build/_gen/tex/wf-lemmas.tex}

@@

\subsection{Variable Ordering}
\input{../_build/_gen/tex/ord-lemmas.tex}

\subsection{Normaliztaion}
\input{../_build/_gen/tex/norm-lemmas.tex}

\subsection{Equivalence}
\input{../_build/_gen/tex/equiv-lemmas.tex}

### 2023-03-28 proofs/tex/norm-lemmas.tex
Prove the UB lemma

\begin{lemma}[Commutativity of normalization and renaming]
  \label{lemma:norm-subst-commute} Normalization of a term commutes with renaming.

  Suppose that $\mu$ is a bijection between two sets of variables
  $\mu : A \leftrightarrow B$. Then
  \begin{itemize}
    \item[$-$] $[[nf([mu]iN)]] = [[ [mu] nf(iN) ]]$
    \item[$+$] $[[nf([mu]iP)]] = [[ [mu] nf(iP) ]]$
  \end{itemize}
  Here equality means alpha-equivalence.
\end{lemma}

\begin{proof}
  Mutual induction on $[[iN]]$ and $[[iP]]$.
  \begin{caseof}
    \item $[[iN]]$ = $[[na]]$ \\
      $[[nf([mu]iN)]] = [[ nf([mu]na) ]] = [[ [mu]na ]] $.
      The latter follows from the fact that $[[ [mu]na ]]$ is a variable, and thus, 
      \ruleref{\ottdruleNrmNVarLabel} is applicable.

      $[[ [mu] nf(iN) ]] = [[ [mu] nf(na) ]] = [[ [mu] na ]] $.
   \item If the type is formed by $[[→]]$, $[[↑]]$, or $[[↓]]$, 
     the required equality follows from the congruence of the normalization and
     substitution, and the induction hypothesis.
     For example, if $[[iN]] = [[iP → iM]]$ then \\
     $\begin{aligned}[t]
        [[nf([mu] iN)]] &= [[ nf([mu] (iP → iM)) ]] \\
                        &= [[ nf([mu]iP → [mu]iM) ]]
                        && \text{By the congruence of substitution} \\
                        &= [[ nf([mu]iP) → nf([mu]iM) ]]
                        && \text{By the congruence of normalization, i.e. \ruleref{\ottdruleNrmArrowLabel}} \\
                        &= [[ [mu]nf(iP) → [mu]nf(iM) ]]
                        && \text{By the induction hypothesis} \\
                        &= [[ [mu](nf(iP) → nf(iM)) ]]
                        && \text{By the congruence of substitution} \\
                        &= [[ [mu]nf(iP → iM) ]]
                        && \text{By the congruence of normalization} \\
                        &= [[ [mu]nf(iN) ]]
      \end{aligned}$ \\
    \item $[[iN]] = [[∀ pas.iM]]$ \label{case:norm-subst-commute} \\
      $\begin{aligned}[t]
          [[ [mu] nf(iN) ]] &= [[ [mu] nf(∀pas.iM)]] \\
                            &= [[ [mu] ∀pas'.nf(iM) ]]
                            && \text {Where $[[ ord {pas} in nf(iM) = pas' ]]$}\\
        \end{aligned}$ \\

      $\begin{aligned}[t]
         [[ nf([mu]iN) ]] &= [[ nf([mu] ∀pas.iM)]] \\
                          &= [[ nf(∀pas.[mu]iM) ]]
                          && \text{Assuming $[[{pas}]] \cap A = \emptyset$
                             and $[[{pas}]] \cap B = \emptyset$}\\
                          &= [[ ∀pbs.nf([mu]iM) ]]
                          && \text {Where $[[ ord {pas} in nf([mu]iM) = pbs ]]$}\\
                          &= [[ ∀pas'.nf([mu]iM) ]]
                          && \text {As $[[pbs]] = [[pas']]$ (see below)}
         \end{aligned}$ \\

    Notice that $[[mu]]$ is free of collisions on $[[{pas}]]$ and $[[fv nf(iM)]]$ because
    \begin{enumerate}[nolistsep]
      \item[(i)] $[[mu]](A \cap [[{pas}]]) \cap [[fv nf(iM)]] = \emptyset \cap [[fv nf(iM)]] = \emptyset$ and
      \item[(ii)] $[[mu]](A \cap [[fv nf(iM)]]) \cap [[{pas}]] \subseteq B \cap
        [[{pas}]] = \emptyset$
    \end{enumerate}
    $\begin{aligned}[t]
       [[pbs]] &= [[ ord {pas} in nf([mu]iM) ]] \\
               &= [[ ord {pas} in [mu]nf(iM) ]]
               && \text{By the induction hypothesis} \\
               &= [[ ord {[mu]pas} in [mu]nf(iM) ]]
               && \text{Since $[[{pas}]] \cap A = \emptyset$} \\
               &= [[ [mu] ord {pas} in nf(iM) ]]
               && \text{by \cref{lemma:distr-mu-ord}} \\
               &= [[ ord {pas} in nf(iM) ]]
               && \text{Since $[[{ord {pas} in nf(iM)}]] \cap A \subseteq [[{pas}]] \cap A = \emptyset$} \\
               &= [[ pas' ]]
     \end{aligned}$ \\

     To show alpha-equivalence of 
     $[[ [mu] ∀pas'.nf(iM) ]]$ and $[[ ∀pas'.nf([mu]iM) ]]$,
     we can assume that $[[{pas'}]] \cap A = \emptyset$, and $[[{pas'}]] \cap B = \emptyset$.
     Then $[[ [mu] ∀pas'.nf(iM) ]] = [[∀pas'.[mu]nf(iM)]] = [[ ∀pas'.nf([mu]iM)
     ]]$, the latter follows from the induction hypothesis.

@@

  \begin{itemize}
    \item[$-$] $[[iN ≈ nf(iN)]]$
    \item[$+$] $[[iP ≈ nf(iP)]]$
  \end{itemize}

\end{lemma}
\begin{proof}
  Mutual induction on $[[nf(iN) = iM]]$ and $[[nf(iP) = iQ]]$.
  Let us consider how this judgment is formed:
  \begin{caseof}
    \item{\nameref{\ottdruleNrmNVarLabel} and \nameref{\ottdruleNrmPVarLabel}}\\ By
      the corresponding equivalence rules.
    \item{\nameref{\ottdruleNrmShiftULabel}, \nameref{\ottdruleNrmShiftDLabel},
        and \nameref{\ottdruleNrmArrowLabel}}\\
      By the induction hypothesis and the corresponding congruent equivalence rules.
    \item{\nameref{\ottdruleNrmForallLabel}}, i.e. $[[nf(∀pas.uN) = ∀pas'.uN']]$ \label{case:norm-soundness:forall}\\
      From the induction hypothesis, we
      know that $[[iN ≈ iN']]$. In particular, by \cref{lemma:equiv-fv}, $[[fv
        iN]] \equiv [[fv iN']]$.
      Then by \cref{lemma:ord-soundness}, $[[{pas'}]]
      \equiv [[{pas} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$, and thus,
      $[[{pas'} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$.

\begin{lemma}[Completeness of quantified normalization]
  Normalization returns the same representative for equivalent types.

@@

>>>>

  \begin{itemize}
    \item[$-$] $[[iN ≈ nf(iN)]]$
    \item[$+$] $[[iP ≈ nf(iP)]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on $[[nf(iN) = iM]]$ and $[[nf(iP) = iQ]]$.
  Let us consider how this judgment is formed:
  \begin{caseof}
    \item{\nameref{\ottdruleNrmNVarLabel} and \nameref{\ottdruleNrmPVarLabel}}\\ By
      the corresponding equivalence rules.
    \item{\nameref{\ottdruleNrmShiftULabel}, \nameref{\ottdruleNrmShiftDLabel},
        and \nameref{\ottdruleNrmArrowLabel}}\\
      By the induction hypothesis and the corresponding congruent equivalence rules.
    \item{\nameref{\ottdruleNrmForallLabel}}, i.e. $[[nf(∀pas.uN) = ∀pas'.uN']]$ \label{case:norm-soundness:forall}\\
      From the induction hypothesis, we
      know that $[[iN ≈ iN']]$. In particular, by \cref{lemma:equiv-fv}, $[[fv
        iN]] \equiv [[fv iN']]$.
      Then by \cref{lemma:ord-soundness}, $[[{pas'}]]
      \equiv [[{pas} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$, and thus,
      $[[{pas'} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$.

\begin{corollary}[Normalization preserves ordering]
  \label{corollary:normalization-ord}
  For any $[[varset]]$,
  \begin{itemize}
  \item[$-$] $[[ord varset in nf(uN)]] = [[ord varset in uM]]$
  \item[$+$] $[[ord varset in nf(uP)]] = [[ord varset in uQ]]$
  \end{itemize}
\end{corollary}
\begin{proof}
  Immediately from \cref{lemma:ord-completeness,lemma:normalization-soundness}.
\end{proof}

\begin{lemma}[Distributivity of normalization over substitution]
  \label{lemma:norm-subst-distr} Normalization of a term distributes over substitution.
  Suppose that $[[Γ2 ⊢ σ : Γ1]]$, i.e. $[[σ]]$ maps variables from $[[Γ1]]$ into types
  taking free variables from $[[Γ2]]$.
  Then
  \begin{itemize}
  \item[$-$] $[[nf([σ]iN)]] = [[ [nf(σ)] nf(iN) ]]$
  \item[$+$] $[[nf([σ]iP)]] = [[ [nf(σ)] nf(iP) ]]$
  \end{itemize}
  where $[[nf(σ)]]$ means pointwise normalization: $[[ [nf(σ)] α⁻]] = [[nf([σ]
  α⁻)]]$. %, and .
\end{lemma}
\begin{proof}
  Mutual induction on $[[iN]]$ and $[[iP]]$.
  \begin{caseof}
    \item $[[iN]]$ = $[[na]]$ \\
      \label{case:norm-subst-distr-var}
      $[[nf([σ]iN)]] = [[ nf([σ]na) ]] = [[ [nf(σ)]na ]] $.

      $[[ [nf(σ)] nf(iN) ]] = [[ [nf(σ)] nf(na) ]] = [[ [nf(σ)] na ]] $.
    \item $[[iP]]$ = $[[pa]]$ \\
      Similar to \cref{case:norm-subst-distr-var}.
   \item If the type is formed by $[[→]]$, $[[↑]]$, or $[[↓]]$, 
     the required equality follows from the congruence of the normalization and
     substitution, and the induction hypothesis.
     For example, if $[[iN]] = [[iP → iM]]$ then \\
     $\begin{aligned}[t]
        [[nf([σ] iN)]] &= [[ nf([σ] (iP → iM)) ]] \\
                        &= [[ nf([σ]iP → [σ]iM) ]]
                        && \text{By the congruence of substitution} \\
                        &= [[ nf([σ]iP) → nf([σ]iM) ]]
                        && \text{By the congruence of normalization, i.e. \ruleref{\ottdruleNrmArrowLabel}} \\
                        &= [[ [nf(σ)]nf(iP) → [nf(σ)]nf(iM) ]]
                        && \text{By the induction hypothesis} \\
                        &= [[ [nf(σ)](nf(iP) → nf(iM)) ]]
                        && \text{By the congruence of substitution} \\
                        &= [[ [nf(σ)]nf(iP → iM) ]]
                        && \text{By the congruence of normalization} \\
                        &= [[ [nf(σ)]nf(iN) ]]
      \end{aligned}$ \\
    \item $[[iN]] = [[∀ pas.iM]]$ \label{case:norm-subst-commute} \\
      $\begin{aligned}[t]
          [[ [nf(σ)] nf(iN) ]] &= [[ [nf(σ)] nf(∀pas.iM)]] \\
                            &= [[ [nf(σ)] ∀pas'.nf(iM) ]]
                            && \text {Where $[[pas']] = [[ ord {pas} in nf(iM)]]
                               = [[ord {pas} in iM]]$
                               (the latter is by
                               \cref{corollary:normalization-ord})}\\
        \end{aligned}$ \\

      $\begin{aligned}[t]
         [[ nf([σ]iN) ]] &= [[ nf([σ] ∀pas.iM)]] \\
                          &= [[ nf(∀pas.[σ]iM) ]]
                          && \text{Assuming $[[{pas} ∩ {Γ1}]] = \emptyset$
                             and $[[{pas} ∩ {Γ2}]] = \emptyset$}\\
                          &= [[ ∀pbs.nf([σ]iM) ]]
                          && \text {Where $[[pbs]] = [[ord {pas} in nf([σ]iM)]]
                             = [[ord {pas} in [σ]iM]]$ (the latter is by
                             \cref{corollary:normalization-ord})}\\
                          &= [[ ∀pas'.nf([σ]iM) ]]
                          && \text{By \cref{lemma:ord-sigma}, $[[pbs]] = [[pas']]$
                             since $[[{pas}]]$ is disjoint with $[[Γ1]]$ and
                             $[[Γ2]]$}\\
                          &= [[ ∀pas'.[nf(σ)]nf(iM) ]]
                          && \text {By the induction hypothesis}\\
         \end{aligned}$ \\

     To show alpha-equivalence of 
     $[[ [nf(σ)] ∀pas'.nf(iM) ]]$ and $[[ ∀pas'.[nf(σ)]nf(iM) ]]$,
     we can assume that $[[{pas'} ∩ {Γ1}]] = \emptyset$, and $[[{pas'} ∩ {Γ2}]]
     = \emptyset$.

@@

\begin{corollary}[Commutativity of normalization and renaming]
  \label{lemma:norm-subst-commute} Normalization of a term commutes with renaming.
  Suppose that $\mu$ is a bijection between two sets of variables
  $\mu : A \leftrightarrow B$. Then
  \begin{itemize}
    \item[$-$] $[[nf([mu]iN)]] = [[ [mu] nf(iN) ]]$
    \item[$+$] $[[nf([mu]iP)]] = [[ [mu] nf(iP) ]]$
  \end{itemize}
\end{corollary}
\begin{proof}
  Immediately from \cref{lemma:norm-subst-distr},
  after noticing that $[[nf(mu)]] = [[mu]]$.
\end{proof}

\begin{lemma}[Completeness of quantified normalization]
  \label{lemma:normalization-completeness}
  Normalization returns the same representative for equivalent types.

@@

\begin{lemma}[Idempotence of normalization]
  \label{lemma:norm-idemp}
  Normalization is idempotent
  \begin{itemize}
  \item[$-$] $[[nf(nf(iN))]] = [[nf(iN)]]$
  \item[$+$] $[[nf(nf(iP))]] = [[nf(iP)]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  By applying \cref{lemma:normalization-completeness} to \cref{lemma:normalization-soundness}.
\end{proof}

\begin{lemma}
  \label{lemma:normal-after-subst}
  The result of a substitution is normalized if and only if the initial type and
  the substitution are normalized.

  Suppose that $[[σ]]$ is a substitution  $[[Γ2 ⊢ σ : Γ1]]$,
  $[[iP]]$ is a positive type ($[[Γ1 ⊢ iP]]$),
  $[[iN]]$ is a negative type ($[[Γ1 ⊢ iN]]$). Then
  \begin{itemize}
  \item[$+$]
    $[[ [σ]iP  ]] \text{ is normal} \iff
    \begin{cases}
      [[σ|fv(iP)]] &\text{is normal}\\
      [[iP]]       &\text{is normal}\\
    \end{cases} $
  \item[$-$]
    $[[ [σ]iN  ]] \text{is normal} \iff
    \begin{cases}
      [[σ|fv(iN)]] &\text{is normal}\\
      [[iN]]       &\text{is normal}\\
    \end{cases} $
  \end{itemize}
\end{lemma}

\begin{proof}
  Mutual induction on $[[Γ1 ⊢ iP]]$ and $[[Γ1 ⊢ iN]]$.
  \begin{caseof}
  \item $[[iN]] = [[na]]$\\
    Then $[[iN]]$ is always normal, and
    the normality of $[[σ|{na}]]$ by the definition means $[[ [σ]na ]]$ is normal.

  \item $[[iN]] = [[iP → iM]]$\\
    \label{case:normal-after-subst-arrow}
    $
    \begin{aligned}[t]
      [[ [σ](iP → iM) ]] \text{ is normal} &\iff [[ [σ]iP → [σ]iM ]] \text{ is normal}
                                           && \text{by the substitution
                                              congruence} \\
                                           &\iff
                                             \begin{cases}
                                             [[ [σ]iP ]] &\text{is normal} \\
                                             [[ [σ]iM ]] &\text{is normal} \\
                                             \end{cases}
                                           && \text{by congruence of normality \ilyam{lemma?}}\\
                                           &\iff
                                             \begin{cases}
                                               [[ iP ]]       &\text{is normal} \\
                                               [[ σ|fv(iP) ]] &\text{is normal} \\
                                               [[ iM ]]       &\text{is normal} \\
                                               [[ σ|fv(iM) ]] &\text{is normal} \\
                                             \end{cases}
                                           && \text{by the induction hypothesis}\\
                                           &\iff
                                             \begin{cases}
                                               [[ iP → iM ]]  &\text{is normal} \\
                                               [[ σ|fv(iP) ∪ fv(iM)]] &\text{is normal} \\
                                             \end{cases}
                                          \iff&&%
                                             \begin{cases}
                                               [[ iP → iM ]]  &\text{is normal} \\
                                               [[ σ|fv(iP→iM)]] &\text{is normal} \\
                                             \end{cases}
    \end{aligned}
    $
  \item $[[iN]] = [[↑iP]]$\\
    By congruence and the inductive hypothesis, similar to \cref{case:normal-after-subst-arrow}
  \item $[[iN]] = [[∀pas.iM]]$\\
    $
    \begin{aligned}[t]
      [[ [σ](∀pas.iM) ]] \text{ is normal} &\iff [[ (∀pas.[σ]iM) ]] \text{ is normal}
                                           && \text{assuming $[[pas]] \cap [[Γ1]] = \emptyset$ and
                                              $[[pas]] \cap [[Γ2]] = \emptyset$} \\
                                           &\iff
                                             \begin{cases}
                                             [[ [σ]iM ]] \text{ is normal} \\
                                             [[ord {pas} in [σ]iM = pas]] \\
                                             \end{cases}
                                           && \text{by the definition of normalization}\\
                                           &\iff
                                             \begin{cases}
                                               [[ [σ]iM ]] \text{ is normal} \\
                                               [[ord {pas} in iM = pas]] \\
                                             \end{cases}
                                           && \text{by \cref{lemma:ord-sigma}}\\
                                           &\iff
                                             \begin{cases}
                                               [[ σ|fv(iM) ]] \text{ is normal} \\
                                               [[ iM ]] \text{ is normal} \\
                                               [[ord {pas} in iM = pas]] \\
                                             \end{cases}
                                           && \text{by the induction hypothesis}\\
                                           &\iff
                                             \begin{cases}
                                               [[ σ|fv(∀pas.iM) ]] \text{ is normal} \\
                                               [[ ∀pas.iM ]] \text{ is normal} \\
                                             \end{cases}
                                           &&
                                              \begin{aligned}
                                              &\text{since $[[fv(∀pas.iM) = fv(iM)]]$;}\\ &\text{by the definition of normalization}
                                              \end{aligned}
    \end{aligned}
    $
  \item $[[iP]] = \dots$\\
    The positive cases are done in the same way as the negative ones.

  \end{caseof}
\end{proof}
